---
title: "regression -  signature cluster"
output: html_document
date: "2026-02-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r include=FALSE}
source('/orfeo/cephfs/scratch/area/lvaleriani/races/ProCESS-examples/assign_signature/plot_res.R')
df <- df %>% 
  select(-patient_id, -sample, -sample_class, -cluster, -RMSE, -cna_caller, -mut_caller) %>% 
  mutate(type = factor(type, levels = c('SBS', 'ID')),
         tool = factor(tool, levels = c('viber', 'pyclonevi', 'mobster_univariate')),
         sig_tool = factor(sig_tool, levels = c('SigProfiler', 'BASCULE')))
```

```{r}
df
```

```{r}
model <- lm(CosineSimilarity ~ . -spn, data = df)
summary(model)
```

-   **Multiple R-squared (**0.1609): This is the most important "reality check." It means your model explains about **16.1%** of the variance in `CosineSimilarity`. While your results are statistically significant, about **84%** of what makes the similarity change is caused by factors *not* in this model (or just random noise).

-   **P-value (**<2.2e−16): Despite the low R-squared, this p-value is extremely small. This means the relationships you found are very likely "real" and not just a fluke in the data.

The variables with three stars (`***`) are your heavy hitters. Here is how they shift the `CosineSimilarity`:

-   **`cluster_processSubclonal` (-0.073):** This is your strongest negative predictor. If a cluster is **Subclonal** (compared to the baseline Clonal), the similarity score drops by about **0.07**. This suggests the model finds subclonal mutations harder to reconstruct accurately.

-   **`toolpyclonevi` (-0.052):** Compared to your baseline tool (likely `viber`), using **PyClone-VI** is associated with a **0.05 drop** in similarity.

-   **`pur` (Purity) (+0.048):** As sample purity increases, similarity increases. This makes sense—cleaner samples usually yield better results.

-   **`cov` (Coverage) (+0.0002):** While significant, the effect is tiny. You’d need a massive increase in coverage to see a major bump in similarity.

-   **`patient_classLow Complexity` (+0.020):** "Low Complexity" patients predictably have higher similarity scores (about **0.02** higher) than the reference class.

-   **`toolmobster_univariate`** and **`sig_toolBASCULE`**: Both have p-values >0.05 (0.18 and 0.20 respectively). In this specific model, these don't have a statistically significant effect on `CosineSimilarity` compared to their baseline counterparts.

```{r}
model_tidied <- tidy(model, conf.int = TRUE) %>%  
  filter(term != "(Intercept)") 

ggplot(model_tidied, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  labs(title = "Effect of Predictors on CosineSimilarity",
       subtitle = "Dots represent the Estimate; bars represent 95% Confidence Intervals",
       x = "Estimate (Change in CosineSimilarity)",
       y = "Predictor Variable") +
  theme_minimal()
```

**Diagnostic Plot - What to look for:**

-   **Residuals vs Fitted:** You want to see a random "cloud" of points. If you see a clear "U" shape or a "fan" shape, a linear model might not be capturing the full complexity of your data.

-   **Normal Q-Q:** You want the points to fall mostly along the diagonal line. If they curve off at the ends, your errors aren't normally distributed.

```{r}
par(mfrow = c(2, 2))
plot(model)
```
